Modelling complex population structure using $F$-statistics and Principal Component Analysis

### 2021-11-09 fstats_pca_lewontin.tex
reorganize discussion

\title{Modelling complex population structure using $F$-statistics and Principal Component Analysis}
\author{Benjamin M Peter}
\begin{document}
	\maketitle
\begin{abstract}
Human  genetic diversity is shaped by our complex history. Population genetic tools to understand this variation can broadly be classified into data-driven methods such as Principal Component Analysis (PCA), and model-based approaches such as $F$-statistics.
Here, I show that these two perspectives are closely related, and I derive explicit connections between the two approaches. I show that $F$-statistics have a simple geometrical interpretation in the context of PCA, and that orthogonal projections are the key concept to establish this link. I illustrate my results on two examples, one of local, and one of global human diversity. In both examples, I find that population structure is sparse, and only a few components contribute to most statistics. Based on these results, I develop novel visualizations that allow for investigating specific hypotheses, checking the assumptions of more sophisticated models. My results extend $F$-statistics to non-discrete populations, moving towards more complete and less biased descriptions of human genetic variation. 
\end{abstract}
\section{Introduction}
As in most species, the genetic diversity of human populations has been influenced by our history and environment over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994,marciniak2017, reich2018a, nielsen2017}. In turn, an important goal of population genetics is to use observed patterns of variation to  investigate and reconstruct the demographic and evolutionary history of our species \citep{schraiber2015, orlando2021}.

The complicated genetic structure observed in present-day human populations \citep{the2015, mallick2016} is caused by the interplay of demographic and evolutionary processes with both discrete and continuous components \citep{pritchard2000, rosenberg2002a, serre2004, rosenberg2005, bradburd2018, reich2018a, peter2020a}.
In particular, populations are expected to slowly differentiate if they are isolated from each other \citep{wahlund1928,cavalli-sforza1975}. In humans, this may be caused because continental-scale geographic distances limit migration, causing a pattern known as isolation-by-distance \citep{slatkin1985}. However, these patterns are usually not uniform, but shaped by geography, particularly barriers  to migration such as mountain ranges, oceans or deserts \citep{cavalli-sforza1994, barbujani1990, rosenberg2005, bradburd2013, peter2020a}. In addition,  major historical population movements such as the out-of-Africa, Austronesian or Bantu expansions lead to more gradual patterns of genetic diversity over space \citep{cavalli-sforza1994, ramachandran2005, novembre2008, stoneking2016, racimo2020}. Local migration between neighboring populations will reduce differentiation, and long-distance migrations \citep{alves2016}, and secondary contact between diverged populations, such as Neandertals and modern humans \citep{green2010} may lead to locally increased diversity \citep{skov2018}.

@@

\subsection{more PCs}
This allows for direct and quantitative comparisons between $F$-statistic-based results and PCA biplots. As PCA is often ran in an early step in data analysis, this may aid in generation of hypotheses that can be more directly evaluated using generative models, typically using a lower number of populations. It also allows reconciling apparent contradictions between $F$-statistics and PCA-plots. In many cases, differences between the two data summaries will be due to variation on higher PCs. In this case, plotting additional PCs, or further subsetting the data to a more local set of populations seems prudent.

@@

\subsubsection{summary}
Thus, while the version of PCA used here differs from that proposed by \cite{patterson2012}, the differences are largely due to conventions, or due to different goals for a particular analysis. Particularly for studies where the description of population structure is a major focus, results might be easier to interpret if  conventions regarding missing data, normalization and estimation of allele frequencies are used consistently between $F$-statistics and PCA.

\subsection{Previous Interpretations}
Previous interpretation of PCA in the context of population genetic models have focused on explicit models and aimed at directly interpreting the PCs in terms of population genetic parameters \citep{cavalli-sforza1975, novembre2008a, francois2021}. My interpretation here is different in that the utility of PCA is to simplify the geometry of the data. One consequence is that the results here are not impacted by sample ascertainment, sample sizes or the number of principal components analyzed, which are common concerns in the interpretation of PCA. However, a very skewed sampling distribution will increases the likelihood that more or different PCs will have to be included in the analysis. From this perspective, one could envision a framework where $F$-statistics are used to decide which samples should be included to obtain a low-dimensional PCA-plot ``representative'' of the data.

\subsection{broader scope}

As $F$-statistics are motivated by trees, they assume that populations are discrete, related as a graph, and that gene flow between populations is rare \citep{patterson2012,harney2021}. However, in many parts of the world, all humans populations are admixed to some degree \citep{pickrell2014}, and in regions such as Europe, genetic diversity is distributed continuously \citep{novembre2008, novembre2008a}. This provides a challenge for interpretation; as most $F_3$ and $F_4$-statistics will indicate departures from treeness. A naive interpretation of the $F$-statistics from my Eurasian example  (Figure \ref{fig:f3}A) would identify a substantial fraction of Europeans as (significantly) admixed between Finnish and Sardinians. In contrast, PCA reveals that the variation in this data set is not due to a single event, and so an arguably better description of the data set is one where Finnish and Sardinians lie on opposite ends of a more gradually structured population.

This trade-off between $F$-statistics, which allow for significance teste tension between

@@

>>>>

\title{Modelling complex population structure using $F$-statistics and Principal Component Analysis}
\author{Benjamin M Peter}
\begin{document}
	\maketitle
\begin{abstract}
Human  genetic diversity is shaped by our complex history. Population genetic tools to understand this variation can broadly be classified into data-driven methods such as Principal Component Analysis (PCA), and model-based approaches such as $F$-statistics.
Here, I show that these two perspectives are closely related, and I derive explicit connections between the two approaches. I show that $F$-statistics have a simple geometrical interpretation in the context of PCA, and that orthogonal projections are the key concept to establish this link. I illustrate my results on two examples, one of local, and one of global human diversity. In both examples, I find that population structure is sparse, and only a few components contribute to most statistics. Based on these results, I develop novel visualizations that allow for investigating specific hypotheses, checking the assumptions of more sophisticated models. My results extend $F$-statistics to non-discrete populations, moving towards more complete and less biased descriptions of human genetic variation. 
\end{abstract}
\section{Introduction}
As in most species, the genetic diversity of human populations has been influenced by our history and environment over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994,marciniak2017, reich2018a, nielsen2017, witt2022}. In turn, an important goal of population genetics is to use observed patterns of variation to  investigate and reconstruct the demographic and evolutionary history of our species \citep{schraiber2015, orlando2021}.

The complicated genetic structure observed in present-day human populations \citep{the2015, mallick2016} is caused by the interplay of demographic and evolutionary processes with both discrete and continuous components \citep{pritchard2000, rosenberg2002a, serre2004, rosenberg2005, bradburd2018, reich2018a, peter2020a, gopalan2022}.
In particular, populations are expected to slowly differentiate if they are isolated from each other \citep{wahlund1928,cavalli-sforza1975}. In humans, this may be caused because continental-scale geographic distances limit migration, causing a pattern known as isolation-by-distance \citep{slatkin1985}. However, these patterns are usually not uniform, but shaped by geography, particularly barriers  to migration such as mountain ranges, oceans or deserts \citep{cavalli-sforza1994, barbujani1990, rosenberg2005, bradburd2013, peter2020a}. In addition,  major historical population movements such as the out-of-Africa, Austronesian or Bantu expansions lead to more gradual patterns of genetic diversity over space \citep{cavalli-sforza1994, ramachandran2005, novembre2008, stoneking2016, racimo2020}. Local migration between neighboring populations will reduce differentiation, and long-distance migrations \citep{alves2016}, and secondary contact between diverged populations, such as Neandertals and modern humans \citep{green2010} may lead to locally increased diversity \citep{gopalan2022}.

@@

\subsection{Previous Interpretations}
Previous interpretation of PCA in the context of population genetic models have focused on explicit models and aimed at directly interpreting the PCs in terms of population genetic parameters \citep{cavalli-sforza1975, novembre2008a, francois2010, francois2021}. My interpretation here is different in that the utility of PCA is to simplify the geometry of the data. One consequence is that the results here are not impacted by sample ascertainment, sample sizes or the number of principal components analyzed, which are common concerns in the interpretation of PCA. However, a very skewed sampling distribution will increases the likelihood that more, or different PCs will have to be included in the analysis.

\subsection{more PCs}
My focus on the geometry of the data allows for direct and quantitative comparisons between $F$-statistic-based results and PCA biplots. As PCA is often ran in an early step in data analysis, this may aid in generation of hypotheses that can be more directly evaluated using generative models, typically using a lower number of populations. It also allows reconciling apparent contradictions between $F$-statistics and PCA-plots. In many cases, differences between the two data summaries will be due to variation on higher PCs. In this case, plotting additional PCs, or further subsetting the data to a more local set of populations seems prudent.

@@

\subsubsection{summary}
The version of PCA used here differs from that proposed by \cite{patterson2006}, and thus some care will be required to directly extend the interpretations developed here to individual-based PCAs. However, the differences are largely due to conventions, and particularly for studies where the description of population structure is a major focus, results might be easier to interpret if conventions regarding missing data, normalization and estimation of allele frequencies are used consistently between $F$-statistics and PCA.

\subsection{The Apportionment of Human Diversity}
Most genetic variation in humans is shared between all of us, but the around 15\% that can be explained by population structure can be leveraged to study our history and diversity in great detail \citep{dobzhansky1972, cavalli-sforza1964, reich2018a}. For some data sets it is possible to predict an individuals' origin at a resolution of a few hundred kilometers \citep{novembre2008, leslie2015}, and direct-to-consumer-genetics companies are using this variation to analyze the genetic data of millions of customers.

However, understanding, conceptualizing and modeling this variation is far from trivial, and misconstrued models of human genetic differentiation have repeatedly been used to justify racist, eugenic and genocidal policies. Lewontin’s landmark 1972 paper on the apportionment of human genetic diversity was one of the first to quantify how little of between-population genetic variation could be attributed to ``racial'' continental-scale groupings. Over the last five decades, this view has been corroborated, refined and extended many times \citep{cann1987, cavalli-sforza1994, barbujani1997, rosenberg2002a}.

From a practical perspective, formulating hypotheses and designing studies in terms of discrete populations with ``uniform'' genetic backgrounds is often sensible, as it enables  e.g. prediction of phenotypes \citep{berg2019, yair2021}, inference of demographic parameters and schematic models of human genetic history \citep{patterson2012}. This is also the case for $F$-statistics, which are motivated by trees. They assume that populations are discrete, related as a graph, and that gene flow between populations is rare \citep{patterson2012,harney2021}.
However, these simplification do come at a cost, both in terms of model violations that may invalidate statistical results, and in terms of deemphasizing people that do not rigidly fall into predefined genetic groups.

 However, in many parts of the world, all humans populations are admixed to some degree \citep{pickrell2014}, and in regions such as Europe, genetic diversity is distributed continuously \citep{novembre2008, novembre2008a}. This provides a challenge for interpretation; as most $F_3$ and $F_4$-statistics will indicate departures from treeness. A naive interpretation of the $F$-statistics from my Eurasian example  (Figure \ref{fig:f3}A) would identify a substantial fraction of Europeans as (significantly) admixed between Finnish and Sardinians. In contrast, PCA reveals that the variation in this data set is not due to a single event, and so an arguably better description of the data set is one where Finnish and Sardinians lie on opposite ends of a more gradually structured population.

@@

\subsubsection{cost of simplificatio}

However, these simplification do come at a cost, both in terms of model violations that may invalidate statistical results, and in terms of excluding people that do not rigidly fall into predefined genetic groups. By framing admixture in the continuous context of PCA, the approach developed here shows one way how this trade-off between interpretability and accuracy can be mitigated.
